Using numerical modelling to investigate temperature and wave exposure as drivers of macroalgal characteristics

### 2019-08-26 morphology_paper.Rmd
minor adjustments to YAML. Cannot get bibtex file to work with pandoc. removing

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers of macroalgal characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  word_document:
    reference_docx: frontiers_template.docx
  pdf_document: 
  html_document:
    df_print: paged
  html_notebook:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
papersize: A4
tables: yes
fontsize: 10pt
---

@@

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera (Bolton 2010), nevertheless form the basis of one of the most productive ecosystems globally (Mann 1973). Kelps generally have a dependence on cool, temperate and arctic seawater temperatures (Bolton 2010; Mohring et al. 2014; Cavanaugh et al. 2011; Dayton 1985), and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres (Steneck et al. 2002). Their size and complex morphology provide a heterogeneous habitat structure (Steneck et al. 2002) that accommodate a multitude of turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates (Mann 1973; Steneck et al. 2002), each depending on a wide suite of ecological services provided by the kelp forests (Gaines & Roughgarden 1987; Paul & Steneck 1993; Levin 1994; Willis & Anderson 2003; Anderson et al. 1997; Dayton 1985). Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in the distribution (Gorman et al. 2013), abundance (Cavanaugh et al. 2011; Dayton et al. 1998), diversity (Wing et al. 2007; Wernberg & Goldberg 2008), composition (Leliaert et al. 2000; Dayton 1985; Norderhaug et al. 2012; Harley et al. 2012), growth (Cousens 1982) and productivity (Pedersen & Nejrup 2012; Dayton et al. 1998; Krumhansl & Scheibling 2012) of kelps.

@@

@@

```{r rda plots, echo=FALSE, fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
ggarrange(lam_rda, rda_eck_deep, rda_eck_shallow, ncol = 1, nrow = 3, label = "AUTO")
lam_rda
rda_eck_deep
rda_eck_shallow
```

Forward selection, assessment of VIF and Pearson correlation coefficients allowed us to retian the most parsimonious descriptors of shallow *E. maxima* morphological characteristics with an adjusted *R*2 of 0.36, explaining 61% of the variation (global permutation test on final model: d.f = 4, *F* = 2.41, *p* = 0.013). The model consisted of one significant axis, RDA1 (p = 0.02) and explained 62% of the variation. Stipe length, stipe mass and stipe circumference were positivley influenced by *max_temp* and *mean_temp* along RDA1, and explained 50% and 55% of the variation respectivley. The remaining morphological characteristics were positivley influenced by by *tp_max* and negativley inlfuenced by *dirw_median* along RDA2 and explained 86% and 94% of the variation respectivley. There was overlapping of clusters according to region with no clear patterns in sites evident.

@@

Past research shows that macroalgal morphological characteristics are largely driven by a wave exposure while others have suggested a complex interaction between various environmental variables. We confirm that wave variables are the main driver of kelp morphological characteristics in both *E. maxima *and *L. pallida* populations around the Cape Peninsula, and differences in shallow and deep populations of *E. maxima* may reflect small differences in wave and temperature variables as drivers of kelp morphological characteristics. The response of kelp morphological characteristics to wave exposure was evident and both species show a tactics based on the magnitde of wave exposure. Strength increasing traits were exibited for lower exposure levels while a 'go-with-the-flow' tactic for moderate levels of wave exposure. Denny and Galord ??? showed that increased flexiblity or a 'go-with-the-flow' tactic has a threshold value, where increased stipe length actaully increase drag under high wave exposure scenarios. Therefore, under high wave exposure scenarios size reducing traits, such as reduced surface area of the overall plant should be favoured instead. This was evident in the results which showed that stipe length decreases under high wave exposure categories. This suggests that how morphological characteristcs manifest themselves is dependent on the magnitude of wave exposure. Furthermore, the results suggest that in the absence of wave exposure, temperatre becomes the dominant morphological driver. This is evident from the redundancy analysis for each species and deep populations of *E. maxima* which show clustering of study sites according to the region; wave variables drive morphological characteristics for kelps from the western region and temperature variables for kelps from the eastern region. Clustering of sites according to region overlapped almost completey for the RDA analysis of shallow *E. maxima* suggesting that for shallow kelp the effect of wave exposure is less variable. Temperature variables were not a significant driver of kelp morphology in this study. Although temperature plays an important role in distribution (Bolton et al. 2012; Miller et al. 2011; Rinde et al. 2014) of kelp and physiological functioning of adults and gametophytes (Mohring et al. 2014; Steneck et al. 2002; Gerard 1997; Bearham et al. 2013; Smale & Moore 2017), there is little evidence that temperature is an important driver of morphological variation. However, we suggest that temperature plays a larger role in determining kelp morphology in areas where wave energy does not exert large amounts of impact.

@@

>>>>

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers of macroalgal characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  word_document:
    reference_docx: frontiers_template.docx
papersize: A4
tables: yes
fontsize: 10pt
bibliography: morphology.bib
---

@@

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera (Bolton 2010), nevertheless form the basis of one of the most productive ecosystems globally (Mann 1973). Kelps generally have a dependence on cool, temperate and arctic seawater temperatures (Bolton 2010; @Mohring et al. 2014; Cavanaugh et al. 2011; Dayton 1985), and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates (Mann 1973; Steneck 2002), each depending on a wide suite of ecological services provided by the kelp forests (Gaines & nRoughgarden 1987; Paul & Steneck 1993; Levin 1994; Willis & Anderson2003; Anderson 1997; Dayton 1985). Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in the distribution (Gorman et al. 2013], abundance (Cavanaugh et al. 2011; Dayton 1998), diversity (Wing 2007; Wernberg & Goldberg 2008), composition (Leliaert et al. 2000; Dayton 1985; Norderhaug et al. 2012; Harley et al. 2012), growth (Cousens 1982) and productivity (Pedersen & Nejrup 2012; Dayton 1998; Krumhansl & Scheibling 2012) of kelps.

@@

```{r rda lam plot, echo=FALSE, fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
lam_rda
```

@@

```{r rda eck plots, echo=FALSE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
ggarrange(rda_eck_deep, rda_eck_shallow, labels = "AUTO", common.legend = TRUE, legend = "bottom", ncol = 1, nrow = 2)
```
Forward selection, assessment of VIF and Pearson correlation coefficients allowed us to retian the most parsimonious descriptors of shallow *E. maxima* morphological characteristics with an adjusted *R*2 of 0.36, explaining 61% of the variation (global permutation test on final model: d.f = 4, *F* = 2.41, *p* = 0.013). The model consisted of one significant axis, RDA1 (p = 0.02) and explained 62% of the variation. Stipe length, stipe mass and stipe circumference were positivley influenced by *max_temp* and *mean_temp* along RDA1, and explained 50% and 55% of the variation respectivley. The remaining morphological characteristics were positivley influenced by by *tp_max* and negativley inlfuenced by *dirw_median* along RDA2 and explained 86% and 94% of the variation respectivley. There was overlapping of clusters according to region with no clear patterns in sites evident.

@@

Past research shows that macroalgal morphological characteristics are largely driven by a wave exposure while others have suggested a complex interaction between various environmental variables. We confirm that wave variables are the main driver of kelp morphological characteristics in both *E. maxima *and *L. pallida* populations around the Cape Peninsula, and differences in shallow and deep populations of *E. maxima* may reflect small differences in wave and temperature variables as drivers of kelp morphological characteristics. The response of kelp morphological characteristics to wave exposure was evident and both species show a tactics based on the magnitde of wave exposure. Strength increasing traits were exibited for lower exposure levels while a 'go-with-the-flow' tactic for moderate levels of wave exposure. Denny and Galord ??? showed that increased flexiblity or a 'go-with-the-flow' tactic has a threshold value, where increased stipe length actaully increase drag under high wave exposure scenarios. Therefore, under high wave exposure scenarios size reducing traits, such as reduced surface area of the overall plant should be favoured instead. This was evident in the results which showed that stipe length decreases under high wave exposure categories. This suggests that how morphological characteristcs manifest themselves is dependent on the magnitude of wave exposure. Furthermore, the results suggest that in the absence of wave exposure, temperatre becomes the dominant morphological driver. This is evident from the redundancy analysis for each species and deep populations of *E. maxima* which show clustering of study sites according to the region; wave variables drive morphological characteristics for kelps from the western region and temperature variables for kelps from the eastern region. Clustering of sites according to region overlapped almost completey for the RDA analysis of shallow *E. maxima* suggesting that for shallow kelp the effect of wave exposure is less variable. Temperature variables were not a significant driver of kelp morphology in this study. Although temperature plays an important role in distribution (Bolton et al. 2012; Miller et al. 2011; Rinde et al. 2014) of kelp and physiological functioning of adults and gametophytes (Mohring et al. 2014; Steneck et al. 2002; Gerard 1997; Bearham et al. 2013; Smale & Moore 2017), there is little evidence that temperature is an important driver of morphological variation. However, we suggest that temperature plays a larger role in determining kelp morphology at sites where wave energy is low enough not to elecit a response in morphological characteristics.

@@

# References
